---
title: "Untitled"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(janitor)
library(plotly)
```

Today we're going to make a reproducible report about volatile oils in dead juniper trees. This is a collaboration  with [YoungLiving Essential Oil, Inc.](https://www.youngliving.com)

First off, we have to load the data set: 

```{r,message=FALSE, warning=FALSE}
df <- read.csv("./Data/Juniper_Oils.csv")
```

**BUT... this data needs some cleaning before we can make it useful. ** It is currently in a wide format, where each compound is in it's own column. 


```{r, message= FALSE, warning=FALSE}
names(df)  <- names(df) %>% make_clean_names()
compounds <- names(df)[11:33]
long <- df %>%
  filter(amplicon == "16S") %>% 
  pivot_longer(compounds, names_to = "compound", values_to= "yield")
```

Now that it's cleaned up a bit, we can plot the compound yields over time since burn... 

```{r, message= FALSE, warning=FALSE}

p1 <- ggplot(long, aes(x=years_since_burn, y=yield)) +
  geom_smooth() +
  geom_point() +
  facet_wrap(~compound, scales =  "free")

```


Here's some summary info about the compounds that were observed: 


```{r, message= FALSE, warning=FALSE, echo=FALSE}
summarized_compound <- long %>% 
  group_by(compound) %>% 
  summarize(MeanConcentration = mean(yield))

knitr :: kable(summarized_compound,row.names = FALSE, digits = 4,caption = "Mean concentration values across entire study")
```

We can also do something to make this plot interactive!

**This uses the "plotly" package** 

```{r, message= FALSE, warning=FALSE}
ggplotly(p1)
```






