---
title: "MoneyBball"
author: "Daines"
date: "4/24/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
library(plotly)
library(dplyr)
library(tidyverse)
library(DT)
```

This project is inspired by Moneyball: The Art of Winning an Unfair Game, a book by Michael Lewis, published in 2003, about the Oakland Athletics baseball team and its general manager Billy Beane. Its focus is the team's analytical, evidence-based, sabermetric approach to assembling a competitive baseball team despite Oakland's small budget. 
(Source: https://en.wikipedia.org/wiki/Moneyball)

### Can this be replicated in basketball?
Could data analytics be applied to NBA player statistics to assemble a competitive team on a small budget?

### Data Sources
+ NBA Salaries - USA Today Sports -  https://hoopshype.com/salaries/
+ NBA Team Salaries - USA Today Sports -  https://hoopshype.com/salaries/
+ NBA Players Statistics - NBA.com - https://www.nba.com/stats/players

```{r, echo=FALSE}

# Data Sources:
# 1) Salary Data
#    USA Today Sports
#    https://hoopshype.com/salaries/
#
# 2) Statistics Data
#    NBA.com
#    https://www.nba.com/stats/players
#
#
# Glossary: (Traditional Stats)
# -------------------------------------------------------
# GP Games Played 
# W Wins 
# L Losses 
# MIN Minutes Played 
# FGM Field Goals Made 
# FGA Field Goals Attempted 
# FG% Field Goal Percentage 
# 3PM 3 Point Field Goals Made 
# 3PA 3 Point Field Goals Attempted 
# 3P% 3 Point Field Goals Percentage 
# FTM Free Throws Made 
# FTA Free Throws Attempted 
# FT% Free Throw Percentage 
# OREB Offensive Rebounds 
# DREB Defensive Rebounds 
# REB Rebounds 
# AST Assists 
# TOV Turnovers 
# STL Steals 
# BLK Blocks 
# PF Personal Fouls 
# FP Fantasy Points 
# DD2 Double doubles 
# TD3 Triple doubles 
# PTS Points 
# +/- Plus Minus
#
# Glossary: (Advanced)
# -------------------------------------------------------
# GP Games Played 
# W Wins 
# L Losses 
# MIN Minutes Played 
# OffRtg Offensive Rating 
# DefRtg Defensive Rating 
# NetRtg Net Rating 
# AST% Assist Percentage 
# AST/TO Assist to Turnover Ratio 
# AST Ratio Assist Ratio 
# OREB% Offensive Rebound Percentage 
# DREB% Defensive Rebound Percentage 
# REB% Rebound Percentage 
# TO Ratio Turnover Ratio 
# eFG% Effective Field Goal Percentage 
# TS% True Shooting Percentage 
# USG% Usage Percentage 
# PACE Pace 
# PIE Player Impact Estimate
#
# Glossary: (Misc)
# -------------------------------------------------------
# GP Games Played 
# W Wins 
# L Losses 
# MIN Minutes Played 
# PTS OFF TO Points Off Turnovers 
# 2nd PTS Second Chance Points 
# FBPs Fast Break Points 
# PITP Points in the Paint Opp 
# PTS OFF TO Opponent Points Off Turnovers 
# Opp 2nd PTS Opponent 2nd Chance Points 
# Opp FBPs Opponent Fast Break Points 
# Opp PITP Opponent Points in the Paint 
# BLK Blocks 
# BLKA Blocked Field Goal Attempts 
# PF Personal Fouls 
# PFD Personal Fouls Drawn
#
# Glossary: (Scoring)
# -------------------------------------------------------
# GP Games Played 
# W Wins 
# L Losses 
# MIN Minutes Played 
# %FGA 2PT Percent of Field Goals Attempted (2 Pointers) 
# %FGA 3PT Percent of Field Goals Attempted (3 Pointers) 
# %PTS 2PT Percent of Points (2 Pointers) 
# %PTS 2PT MR Percent of Points (Mid-Range) 
# %PTS 3PT Percent of Points (3 Pointers) 
# %PTS FBPs Percent of Points (Fast Break Points) 
# %PTS FT Percent of Points (Free Throws) 
# %PTS OffTO Percent of Points (Off Turnovers) 
# %PTS PITP Percent of Points (Points in the Paint) 
# 2FGM %AST Percent of 2 Point Field Goals Made Assisted 
# 2FGM %UAST Percent of 2 Point Field Goals Made Unassisted 
# 3FGM %AST Percent of 3 Point Field Goals Made Assisted 
# 3FGM %UAST Percent of 3 Point Field Goals Made Unassisted 
# FGM %AST Percent of Point Field Goals Made Assisted 
# FGM %UAST Percent of Point Field Goals Made Unassisted
#
# Glossary: (Defense)
# -------------------------------------------------------
# GP Games Played 
# W Wins 
# L Losses 
# MIN Minutes Played 
# DEF RTG Defensive Rating 
# DREB Defensive Rebounds 
# DREB% Defensive Rebound Percentage 
# % DREB Percent of Team's Defensive Rebounds 
# STL Steals 
# STL% Percent of Team's Steals 
# BLK Blocks 
# %BLK Percent of Team's Blocks 
# OPP PTS OFF TOV Opponent Points Off Turnovers 
# OPP PTS 2ND CHANCE Opponent 2nd Chance Points 
# OPP PTS FB Opponent Fast Break Points 
# OPP PTS PAINT Opponent Points in the Paint 
# DEF WS Defensive Win Shares
#
# Glossary: (Estimated Advanced)
# -------------------------------------------------------
# GP Games Played 
# W Wins 
# L Losses 
# MIN Minutes Played 
# Est. OffRtg Estimated Offensive Rating 
# Est. DefRtg Estimated Defensive Rating 
# Est. NetRtg Estimated Net Rating 
# Est. AST Ratio Estimated Assist Ratio 
# Est. OREB% Estimated Offensive Rebound Percentage 
# Est. DREB% Estimated Defensive Rebound Percentage 
# Est. REB% Estimated Rebound Percentage 
# Est. TO Ratio Estimated Turnover Ratio 
# Est. USG% Estimated Usage Percentage 
# Est. PACE Estimated Pace

```

### Loading Salary Data

#### Cleaning Data

+ Load from csv file
+ Remove unneeded columns
+ Rename columns
+ Remove header row from source data
+ Convert salary ($,) from chr to numeric
+ Sort by player name (to later combine with statistics)

```{r, echo=FALSE}
# Load NBA player salary data
ps <- read.csv("nba-salaries.csv")

#dim(ps)
#str(ps)
#colnames(ps)

# select only the columns we need: Player name and 20/21 salary
ps2 <- ps %>% dplyr::select(2:3) 

#head(ps2)
#colnames(ps2)

# change column names from defaults
ps2 <- ps2 %>%
  dplyr::rename(Player = Column.2,
                Salary.20.21 = Column.3
              )

#colnames(ps2)
# remove header row
ps2 <- ps2[-c(1),]
#head(sal_df2)

# convert salary from chr with $ and ,'s to numeric
ps2$Salary.20.21 <- as.numeric(gsub('[$,]', '', ps2$Salary.20.21))

# sort by player name
ps2 <- ps2 %>%
  arrange(Player)

#head(ps2)
```

Here is a boxplot of player salaries:

```{r, message=FALSE,warning=FALSE,echo=FALSE,fig.align='center', fig.width=2, fig.height=3}

ggplot(ps2, aes(y=Salary.20.21/1000000)) +
  geom_boxplot() +
  theme_minimal() +
  ylab("Player Salaries ($M)")

```

Search for your favorite NBA player to see how much money they are being paid per yer. Some of these might surprise you. 

```{r, echo=FALSE, message=FALSE, fig.align='center', fig.width=2, fig.height=3 }
datatable(ps2, rownames = FALSE, filter="top", options = list(pageLength = 5, scrollX=T) )
```



```{r, echo=FALSE}
#-------------------------------------------------------
# Load NBA team salary data
ts <- read.csv("nba-team-salaries.csv")

#dim(ts)
#str(ts)
#colnames(ts)
#head(ts)

# select only the columns we need: Player name and 20/21 salary
ts2 <- ts %>% dplyr::select(2:3) 

#head(ts2)
#colnames(ts2)

# change column names from defaults
ts2 <- ts2 %>%
  dplyr::rename(Team.Name = Column.2,
                Salary.20.21 = Column.3
              )

# convert salary from chr with $ and ,'s to numeric
ts2$Salary.20.21 <- as.numeric(gsub('[$,]', '', ts2$Salary.20.21))

# sort by team name
ts2 <- ts2 %>%
  arrange(Team.Name)

#head(ts2)

#-------------------------------------------------------
# Load NBA team record data
tr <- read.csv("nba-team-record.csv")

#dim(tr)
#str(tr)
#colnames(tr)
#head(tr)

# select only the columns we need: Player name and 20/21 salary
tr2 <- tr %>% dplyr::select(1:3) 

#head(tr2)
#colnames(tr2)


# change column names from defaults
tr2 <- tr2 %>%
  dplyr::rename(Team.Name = Column.1,
                Win = Column.2,
                Loss = Column.3
              )
tr2 <- tr2 %>% 
# remove conference standing digit(s) from name
  mutate(Team.Name = as.character(gsub("^\\d*","",Team.Name))) %>%
# remove team name
  mutate(Team.Name = as.character(gsub(" .*","",Team.Name)))

# sort by city
tr2 <- tr2 %>%
  arrange(Team.Name)

tr2$Salary <- ts2$Salary.20.21[match(tr2$Team.Name, ts2$Team.Name)]
#head(tr2)

```

This shows boxplot of team salaries:

```{r, message=FALSE,warning=FALSE,echo=FALSE,fig.align='center', fig.width=2, fig.height=3}

ggplot(ts2, aes(y=Salary.20.21/1000000)) +
  geom_boxplot(aes()) +
  theme_minimal() +
  ylab("20-21 NBA Team Salaries ($M)")
  
```

As depicted in boxplot, the salaries are grouped around $130M. Yet, for approximately the same team salary there is a wide range of wins (from <15 to >40)!

```{r, message=FALSE,warning=FALSE,echo=FALSE,fig.align='center', fig.width=6, fig.height=4}

ggplot(tr2, mapping = aes(x=Win, y=Salary/1000000)) +
  geom_point(aes(size=Salary/1000000, color=Salary/1000000), show.legend=FALSE) +
  theme_minimal() +
  xlab("Wins") +
  ylab("20-21 NBA Team Salaries ($M)")
  
```

### Loading Player Statistics Data

#### Player Statistics Data

```{r, echo=FALSE}
# Load NBA player statistics data
st1_df <- read.csv("NBA-player-stats-traditional.csv")

# sort by player name
st1_df2 <- st1_df %>%
  arrange(Player)

colnames(st1_df2)[30] <- "P.M"

```

```{r, echo=FALSE}
# Load NBA player statistics - advanced - data
st2_df <- read.csv("nba-stats-advanced.csv")
#head(st2_df)

# select only the columns we need: Player name and 20/21 salary
st2_df2 <- st2_df %>% dplyr::select(2:24) 
#head(st2_df2)

# change column names from defaults
st2_df2 <- st2_df2 %>%
  dplyr::rename(Player = Column.2,
                Adv.Team = Column.3,
                Adv.Age = Column.4,
                Adv.GP = Column.5,
                Adv.W = Column.6,
                Adv.L = Column.7,
                Adv.Min = Column.8,
                Adv.OffRtg = Column.9,
                Adv.DefRtg = Column.10,
                Adv.NetRtg = Column.11,
                Adv.Ast.P = Column.12,
                Adv.Ast.TO = Column.13,
                Adv.AstR = Column.14,
                Adv.OReb.P = Column.15,
                Adv.DReb.P = Column.16,
                Adv.Reb.P = Column.17,
                Adv.TOR = Column.18,
                Adv.EFG.P = Column.19,
                Adv.TS.P = Column.20,
                Adv.USG.P = Column.21,
                Adv.Pace = Column.22,
                Adv.PIE = Column.23,
                Adv.Poss = Column.24
              )

st2_df2 <- st2_df2 %>%
  dplyr::select(!(Adv.Team:Adv.Min))

st3_df1 <- dplyr::inner_join(st1_df2, st2_df2, by="Player")
st3_df2 <- dplyr::inner_join(st3_df1, ps2, by="Player")

paste0(ncol(st3_df2)-1," Statistics for ",nrow(st3_df1)," Players have been loaded.")

#head(st3_df2)
```

#### Analyzing Net Rating vs. Salary

```{r, message=FALSE,warning=FALSE,echo=FALSE,fig.align='center', fig.width=6, fig.height=4}

ggplot(st3_df2, aes(x=Adv.NetRtg, y=Salary.20.21/1000000)) +
  geom_point(aes(), show.legend=FALSE) +
  theme_minimal() +
  xlab("Net Rating") +
  ylab("Salary ($M)")
  
```


#### Analyzing +/- vs. Salary

```{r, message=FALSE,warning=FALSE,echo=FALSE,fig.align='center', fig.width=6, fig.height=4}

ggplot(st3_df2, aes(x=P.M, y=Salary.20.21/1000000)) +
  geom_point(aes(color=factor(MIN), size=factor(MIN), alpha = 0.5), 
             show.legend = FALSE) +
  theme_minimal() +
  scale_color_hue(direction = -1) +
  xlab("Plus/Minus") +
  ylab("Salary ($M)")
  
```

#### Let's filter out players
+ With few minutes (we need players!)
+ And Plus/Minus rating < 4 (we need good players)
+ With salaries over $20M (we need affordable players)

```{r, message=FALSE,warning=FALSE,echo=FALSE,fig.align='center', fig.width=6, fig.height=4}

st3_df3 <- st3_df2 %>% 
  filter(MIN > 20) %>%
  filter(P.M > 4) %>%
  filter(Salary.20.21 < 15000000)

paste0("Now we have ",nrow(st3_df3)," players to analyze.")

ggplot(st3_df3, aes(x=P.M, y=Salary.20.21/1000000)) +
  geom_point(aes(color=factor(MIN), size=factor(MIN), alpha = 0.5), 
             show.legend = FALSE) +
  theme_minimal() +
  scale_color_hue(direction = -1) +
  xlab("Plus/Minus") +
  ylab("Salary ($M)")
  
```
```{r, message=FALSE,warning=FALSE,echo=FALSE,fig.align='center', fig.width=6, fig.height=4}


st4_df1 <- as.data.frame(cbind(st3_df3$Player, 
                            st3_df3$P.M, 
                            st3_df3$Salary))

colnames(st4_df1) <- c("Player", "P.M", "Salary")

st4_df1$Salary <- as.numeric(st4_df1$Salary)

st4_df1 <- st4_df1 %>%
  arrange(order(Salary))


st4_df1

new_salary <- sum(st4_df1$Salary)
new_salary

```


